Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-11-14 _tex/index.tex
Built site for gh-pages

In \emph{comparative judgment} (CJ) studies, judges assess a specific
trait or competence across various stimuli by performing pairwise
comparisons \citep{Thurstone_1927, Pollitt_2004, Pollitt_2012a}. Each
comparison produces a dichotomous outcome, indicating which stimulus is
perceived to manifest a higher trait level. For example, when assessing
text quality, judges compare pairs of written texts (the stimuli) to
determine the relative quality each text exhibit (the trait)
\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.

@@

Nevertheless, despite the growing number of CJ studies, unsystematic and
fragmented research approaches in the literature have left several
critical issues unaddressed. This research primarily focuses on three:
the apparent disconnect between CJ's measurement and structural model,
the over-reliance on the assumptions of Thurstone's Case 5
\citeyearpar{Thurstone_1927} in CJ's measurement model, and the unclear
role of comparison algorithms on the method's reliability and validity.
The following sections will discuss each of these issues in detail,
followed by the introduction of a theoretical model and its statistical
translation, which aims to address all three concerns simultaneously.

@@

@@

In a typical CJ study, the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959} serves as the measurement model
for CJ \citep{Andrich_1978, Bramley_2008}. A measurement model specifies
how manifest variables contribute to the estimation of latent variables
\citep{Everitt_et_al_2010}. For example, when evaluating text quality,
researchers use the BTL model to process the dichotomous outcomes
resulting from the pairwise comparisons (the manifest variables) to
estimate scores that reflect the underlying quality level of the texts
(the latent variable)
\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.

@@

To mitigate these risks, principles from Structural Equation Modeling
(SEM) \citep[pp.~138]{Hoyle_et_al_2023} and Item Response Theory (IRT)
\citetext{\citealp[chap.~6]{Fox_2010}; \citealp[chap.~24]{vanderLinden_et_al_2017_I}}
recommend conducting these analyses and tests within a structural model.
A structural model specifies how different manifest or latent variables
influence the latent variable of interest \citep{Everitt_et_al_2010}.
This approach allows analyses that can account for both the scores and
their uncertainties simultaneously, rather than treating them as
separate elements. Therefore, an integrated approach that combines CJ's
measurement and structural models can offer significant advantages.

\subsection{The assumptions of Case 5 and the measurement
model}\label{the-assumptions-of-case-5-and-the-measurement-model}

>>>>

In \emph{comparative judgment} (CJ) studies, judges assess a specific
trait or attribute across various stimuli by performing pairwise
comparisons \citep{Thurstone_1927, Pollitt_2004, Pollitt_2012a}. Each
comparison produces a dichotomous outcome, indicating which stimulus is
perceived to exhibit a higher trait level. For example, when assessing
text quality, judges compare pairs of written texts (the stimuli) to
determine the relative quality each text exhibit (the trait)
\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.

@@

Nevertheless, despite the increasing number of CJ studies, unsystematic
and fragmented research approaches have left several critical issues
unaddressed. This research primarily focuses on three: the over-reliance
on Thurstone's Case V assumptions in the statistical analysis of CJ
data, the apparent disconnect between CJ's trait measurement and
hypothesis testing, and the unclear role and influence of comparison
algorithms on the method's reliability and validity. The following
sections will discuss each of these issues in detail, followed by the
introduction of a theoretical model and its statistical translation,
which aims to address all three concerns simultaneously.

@@

\subsection{The Case V and the statistical analysis of CJ
data}\label{the-case-v-and-the-statistical-analysis-of-cj-data}

In its most general form, Thurstone's theory
\citeyearpar{Thurstone_1927} posits that when a judge compares two
stimuli, the resulting dichotomous outcome depends on two main factors:
the discriminal process and the law of comparative judgment. The
\emph{discriminal process} refers to the psychological effect each
stimulus has on the judges, or more simply stated, the judges'
perception of the trait level of each stimulus. Thurstone assumes that
the discriminal process for each stimulus follows a Normal distribution.
In this distribution, the mode (mean), known as the \emph{modal
discriminal process}, represents the position of the stimulus on the
trait continuum, while the dispersion, known as the \emph{discriminal
dispersion}, reflects variability in the perceived trait level of the
stimulus. Figure~\ref{fig-discriminal_process} shows a visual depiction
of the discriminal process of two stimuli (objects).

\begin{figure}

\centering{

\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/figures/discriminal_process.png}

}

\caption{\label{fig-discriminal_process}Discriminal processes of two
stimuli (objects). Extracted from \citet[pp.~249]{Bramley_2008}}

\end{figure}%

However, since the discriminal mode and dispersion of a single stimulus
are not directly observable without comparing it to other stimuli, the
\emph{law of comparative judgment} becomes essential. This law asserts
that when assessing a specific trait by comparing two stimuli, the
stimulus positioned further along the continuum is perceived as having a
higher level of that trait. Thus, the observed dichotomous outcome is
determined by the distribution of the difference between the discriminal
processes of the two stimuli, called the \emph{discriminal difference}.
Figure~\ref{fig-comparative_judgment} shows a visual depiction of the
discriminal difference of two stimuli (objects).

\begin{figure}

\centering{

\includegraphics[width=0.65\linewidth,height=\textheight,keepaspectratio]{./images/figures/comparative_judgment.png}

}

\caption{\label{fig-comparative_judgment}Discriminal difference between
two stimuli (objects). Extracted from \citet[pp.~251]{Bramley_2008}}

\end{figure}%

Thurstone's general form of the theory primarily applies to pairwise
comparisons of stimuli made by a single judge
\citep[pp.~267]{Thurstone_1927}. Therefore, to enhance the theory's
practical applicability, he developed five distinct cases, each
incorporating a series of increasingly simplifying assumptions.
Figure~\ref{fig-thurstone_cases} outlines each case along with their key
assumptions. For a detailed explanation of this theoretical progression,
see \citet{Thurstone_1927} and \citet[pp.~248-253]{Bramley_2008}.

\begin{figure}

\centering{

\includegraphics[width=0.85\linewidth,height=\textheight,keepaspectratio]{./images/tables/thurstone_cases.png}

}

\caption{\label{fig-thurstone_cases}Thurstones cases and asumptions}

\end{figure}%

Surprisingly, Case V is the most commonly used case in the CJ
literature, despite incorporating the largest number of simplifying
assumptions \citep[pp.~253]{Bramley_2008}. This widespread adoption is
primarily due to the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959}, which offers a simpler
statistical representation of the case. The BTL model closely aligns
with Case V's assumptions, with one key difference: the distribution of
the discriminal process. While Case V assumes a Normal distribution, the
BTL model uses the more mathematically tractable logistic distribution
\citep[pp.~254]{Bramley_2008} (see Figure~\ref{fig-thurstone_cases}).
However, this adjustment does not significantly affect the model's
estimation or interpretation, as the logistic and normal distributions
are related to each other by a factor of approximately \(1.7\)
\citep[pp.~16]{vanderLinden_et_al_2017_I}.

However, because Case V was originally developed to provide a ``rather
coarse scaling'' of traits \citep[pp.~269]{Thurstone_1927}, prioritizing
statistical ease of use over precision in trait measurement, two of its
assumptions may be causing more problems than expected: the assumption
of zero correlation between the stimuli, and the assumption of equal
dispersion between stimuli.

@@

As outlined in the previous section, the BTL model typically functions
as the measurement model in a CJ study
\citep{Andrich_1978, Bramley_2008}. A measurement model specifies how
manifest variables contribute to the estimation of latent variables
\citep{Everitt_et_al_2010}. For example, when evaluating text quality,
researchers use the BTL model to process the dichotomous outcomes
resulting from the pairwise comparisons (the manifest variables) to
estimate scores that reflect the underlying quality level of the texts
(the latent variable)
\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.

@@

To mitigate these risks, principles from Structural Equation Modeling
(SEM) \citep[pp.~138]{Hoyle_et_al_2023} and Item Response Theory (IRT)
\citetext{\citealp[chap.~6]{Fox_2010}; \citealp[chap.~24]{vanderLinden_et_al_2017_I}}
recommend conducting these analyses and tests within a structural model.
A structural model specifies how different manifest or latent variables
influence the latent variable of interest \citep{Everitt_et_al_2010}.
This approach allows analyses that can account for both the BTL scores
and their uncertainties simultaneously, rather than treating them as
separate elements. Therefore, an integrated approach that combines CJ's
measurement and structural models can offer significant advantages.
